Homework 2 - Part 2
========================================================
Xiyang Dai
--------------------------------------------------------

```{r}
library(TxDb.Dmelanogaster.UCSC.dm3.ensGene)
txdb = TxDb.Dmelanogaster.UCSC.dm3.ensGene
theGene = "FBgn0000008"

exonsByGene = exonsBy(txdb,by="gene")
theRegions = disjoin(exonsByGene[[theGene]])

txsByGene = transcriptsBy(txdb)
theTxsIds = values(txsByGene[[theGene]])$tx_id

exonsByTranscript = exonsBy(txdb,by="tx")
theTxs = exonsByTranscript[theTxsIds]
```

## Question 1 Write code to create a matrix C with number of rows equal to the length of object theRegions and number of columns equal to the length of object theTxs, defined as follows:
```{r}
C = matrix(nrow=length(theRegions), ncol=length(theTxs));
for (i in 1:length(theTxs)){
  C[,i] = countOverlaps(theRegions, theTxs[i]@unlistData)
}
print(C)
```

```{r}
library(ShortRead)
# the entire genomic region overlapping the disjoint gene regions
theRegion = range(theRegions)

# load reads in genomic region (extended by 100,000 bases to either side)
theBigRegion = resize(theRegion,width=width(theRegion)+2e5,fix="center")

# there is a mismatch in chromosome names between the transcript annotation we're using
# and the aligned read data, so let's reconcile them
theBigRegion = keepSeqlevels(theBigRegion,"chr2R")
theBigRegion = renameSeqlevels(theBigRegion, c("chr2R"="2R"))

# now load aligned reads for this region 
# assuming the BAM file is in the current directory
aln = readGAlignments("thefile_sorted.bam", param=ScanBamParam(which=theBigRegion))

# change the chromosome names in the aligned reads object
# to match the transcript annotation
chrNames = seqlevels(aln)
renameVector = paste("chr",chrNames,sep="")
names(renameVector) = chrNames
aln = renameSeqlevels(aln,renameVector)

# there is a slight disagreement on the chromosome lengths between the reference used to align
# reads and the transcript annotation. Let's drop the chrU and chrUextra chromosomes
namesToKeep = seqlevels(aln)[!seqlevels(aln) %in% c("chrU","chrUextra")]
aln = keepSeqlevels(aln, namesToKeep)
theRegions = keepSeqlevels(theRegions, namesToKeep)

# last thing, strand is meaningless for this RNA-seq dataset, so let's make all the overlap
# computation functions ignore it by setting strand to "*"
strand(aln) = "*"
```

## Question 2 Are all three transcripts for this gene expressed? (Answer yes/no/can't tell for each one, and why).
```{r}
olaps = summarizeOverlaps(theRegions, aln)
counts = assay(olaps)[,1]
print(counts)
```

From the counts, it is hard to tell the exact expressed isoforms. Since, the first number of extons is 0 (this is unique for the first isoform), we guess the first isoform is not expressed. The second and third are expressed. 

## Question 3 How is a_ij defined in this model?

In the paper, a is defined as:
\[
  a = lwC
\]
where l is the length of exons and wis the total number of reads.
```{r}
w = length(aln)
l = theRegions@ranges@width
A = l*w*C;
print(A)
```

## Question 4 Write down the optimization problem to solve in terms of matrix A computed above. Use θ as the parameters to estimate.

First, we need to model the likilihood function, given observed exton counts 
x_1, x_2, ..., x_n, we want to model
\[
P(x_1,x_2,...,x_n|\theta) = \prod_{i=1}^{n} \frac{e^{-\sum_j a_{ij}\theta_j}(-\sum_j a_{ij}\theta_j)^{x_n}}{!x_n}
\]
Then, we use MLE to optimize this problem
\[
\arg\max_{\theta} \log(P(x_1,x_2,...,x_n|\theta)) \propto \arg\max_{\theta} \log \sum_{i=1}^{n} (\sum_j a_{ij}\theta_j) -\sum_j a_{ij}\theta_j
\]

## Question 5 Write function loglik <- function(theta, counts, A){...} that takes a vector of parameters theta as argument and computes the log-likelihood of the count data using matrix A given current set of parameters theta.
```{r}
loglik = function(theta, counts, A){
  result = 0
  for (i in 1:length(counts)){
    tmp = sum(A[i,]*theta)
    result = result + counts[i]*log(tmp) - tmp
  }
  return(-result)
}

# loglik2 = function(theta, counts, A){
#   - (t(counts)%*%log(A%*%theta) - sum(A%*%theta))
# }
```

## Question 6 Write function loglikGrad <- function(theta, counts, A) {...} that takes a vector parameters theta as argument and computes the gradient of the log-likelihood at current value of parameters theta.
```{r}
loglikGrad = function(theta, counts, A){
  result = theta*0
  for (j in 1:length(theta)){
    result_j = 0
    for (i in 1:length(counts)){
      result_j = result_j + counts[i]*(A[i,j]/(sum(A[i,]*theta))) - A[i,j]
    }
    result[j] = result_j
  }
  return(-result)
}

# loglikGrad2 = function(theta, counts, A){
#   -(t(counts)%*%(A/colSums(A*theta)) - colSums(A))
# }
```

## Question 7 Write a function estimateTheta <- function(counts, A) {..} that uses the loglik and loglikGrad functions to compute the Maximum Likelihood Estimate of theta. 
```{r}
estimateTheta = function(counts, A){
  ntheta = length(A[1,])
  # initialize \theta s.t. \sum_j \theta_j = 1
  theta = A[1,]*0 + 1/ntheta
  constrOptim(theta, loglik, loglikGrad, ui=matrix(c(1,0,0,0,1,0,0,0,1),3,3), ci=c(0,0,0), counts=counts, A=A)
}

theta = estimateTheta(counts,A)
print(theta$par)
```
### Discussion
From the optimized result we can see all three isoforms are expressed. And from the quantitive level, we can tell that the first and the second isoforms express more then the third one. 